Balsa: A Compact Line Notation Based on SMILES

### 2022-09-30 manuscript.tex
Fixes inconisistency between grammar and manuscript.
Makes several line edits.

\begin{abstract}
Despite its widespread use, Simplified Molecular Input Line Entry System (SMILES) remains underspecified. The lack of a detailed specification encourages improvisation by software developers, complicates data standardization efforts, and undermines extension development. To address these problems this work proposes Balsa, a language subset of SMILES. A formal, machine-readable grammar defines Balsa's syntax. Semantics are described at three levels of resolution: constitution; configuration; and conformation. To estimate the compatibility of Balsa strings with SMILES software, all differences between the two languages were identified. To the extent that the SMILES documentation is unambiguous, every difference was consistent with a language subset relationship.
\end{abstract}

@@

Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988~\cite{weininger:1988}. As a line notation~\cite{wiswesser:1968}, SMILES represents molecules as single line character sequences, or strings. SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel~\cite{oBoyle:2011}; RDKit~\cite{rdkit}; Chemistry Development Kit~\cite{steinbeck:2003}; JChem Base~\cite{jchemBase}; the Daylight Toolkit~\cite{daylightToolkit}; and OEChem TK~\cite{oetk}. SMILES encodings can be found in many public-facing databases, including: PubChem~\cite{kim:2016}; ChEBI~\cite{degtyarenko:2007}; ZINC~\cite{irwin:2005}; ChEMBL~\cite{gaulton:2012}; and Wikipedia~\cite{wikipediaCAS}. Raw SMILES strings have been used extensively in machine learning applications~\cite{sousa:2021}. Algorithmic selection of a single SMILES encoding has been used for unique molecular identification~\cite{weininger:1989,oBoyle:2012}. SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES~\cite{hanson:2016}, CurleySMILES~\cite{drefahl:2011}, BigSMILES~\cite{lin:2019}, and CXSMILES~\cite{cxsmiles}.

@@

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2003}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. (\enquote{Daylight}), recapitulates published material~\cite{daylightTheory}. The lack of other authoritative sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

@@

More broadly, this situation undermines ongoing data integrity efforts underway in other scientific disciplines. As an example, the FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. The lack of a precise specification means that chemical data sets based on SMILES will suffer in all four areas.

@@

Multiple molecules are supported through \textit{gaps}. A gap is an edge devoid of bonding. Gaps can be used whenever two or more disconnected molecules need to be encoded within the same molecular tree. Specific applications include ionic bonding and non-ionic complexes. However, no bonding relationship whatsoever is implied by the presence of a gap. In particular, the zero-order bond interpretation is excluded \cite{clark:2011}.

@@

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen atom and its bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. However, not all atoms are eligible for this treatment (see: \nameref{computing-implicit-hydrogen-count}).

@@

\begin{table}
\caption{Atom Parity Variants.}
\centering
\begin{tabular}{l l}
    \hline
    Variant & Description  \\
    \hline
    \ttt{Clockwise} & Configuration runs with node order. \\
    \ttt{Counterclockwise} & Configuration runs against node order. \\
    \hline
\end{tabular}
\label{table:atom-parity-variants}
\end{table}

The \ttt{edges} ("edges") attribute is a possibly empty ordered list of edges associated with an atom. As noted in the previous section, an edge may represent either a bond or a gap (Table~\ref{table:edge-variants}). The mere presence of an edge does not guarantee a bonding relationship.

@@

A bond is composed of four attributes (Table~\ref{table:bond-attributes}). The \ttt{order} attribute (\enquote{order}) represents the concept of formal bond order, or the number of bonding electrons divided by two. A bond's order may only assume the values one, two, or three. The \ttt{elided} attribute (\enquote{elided}) allows a bond to be omitted during serialization and participate in extended bonding. The \ttt{direction} attribute (\enquote{direction}) is used with other information to conformationally restrict a bond. The two allowed variants for \ttt{Direction}, \ttt{Up} and \ttt{Down} (Table~\ref{table:bond-direction-variants}), will be described in detail later (see: \nameref{conformation}). The \ttt{target} attribute identifies the child node at which the bond is directed (Table~\ref{table:target-variants}).

\begin{table}
\caption{Bond Attributes.}
\centering
\begin{tabular}{l p{3.6cm} l l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{order} & Formal bond order & \ttt{1..4} & 1 \\
    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
    \ttt{direction} & Partial conformational parity & \ttt{\{Direction\}} & \ttt{None} \\
    \ttt{target} & An atom or bridge node & \ttt{Target} & - \\
    \hline
\end{tabular}
\label{table:bond-attributes}
\end{table}

@@

The atom and bond attributes defined here will be referenced throughout this paper as a way to connect syntax and semantics. Implementations may, of course, use any suitable internal data model. They must, however, ensure that such a model is consistent with the one provided here.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME, Figure~\ref{fig:dime}). DIME arises when a given molecule can be represented by two forms, each differing from the other only in double bond placement. DIME is sometimes referred to as \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

Selection is nothing more than an alternative to explicit bond order encoding. No other properties whatsoever are conferred to selected atoms. The members of a DS may be assigned the labels \enquote{aromatic,} \enquote{antiaromatic,} \enquote{non-aromatic,} or any number of other designations by downstream applications. Two encodings differing only in the presence of a DS should be considered equivalent. Balsa itself provides no guidance at all about whether or not selectable atoms should be selected, although individuals and organizations may do so.

@@

To support implicit hydrogen counting, Balsa uses the concept of \textit{valence}. Valence is a non-negative integer computed as the sum of bond orders at a given atom. Single and elided bonds contribute one to the tally, double bonds two, and triple bonds three. Each virtual hydrogen contributes one. For example, the valence of a methyl carbon having three virtual hydrogens and a hydrogen neighbor is four. The valence of an acetaldehyde oxygen atom is two. And so on.

@@

Restricted rotation about a bond leads to conformational isomers (\enquote{conformers}). When the bond in question is a double bond, two conformers result. They can be distinguished through \textit{partial parity bonds} (PPBs). As the name implies, a PPB encodes some of the parity information characterizing a conformationally restricted double bond. Reconstruction of the full parity requires the double bond itself and at least one flanking PPB at each terminal.

@@

The configuration about an atom can specified through its \ttt{parity} attribute. This attribute, together with a general convention, determine the relative three-dimensional positioning of substituents about a central atom (Figure~\ref{fig:atom-parity}). First, the parent of the central atom is identified. Then the frame of reference is shifted by sighting down the bond from the central atom's parent to the central atom. An equilateral triangle is then inscribed about the central atom. The bonds from the central atom are then iterated in order. If the parity of the central atom equals \ttt{Clockwise}, children placed around the triangle in clockwise order. Otherwise the children are placed in counterclockwise order.

@@

The convention is modified for atoms bearing one virtual hydrogen and three atomic substituents (Figure~\ref{fig:atom-parity-suppressed-hydrogen}). The implied bond to the virtual hydrogen is considered to precede the others. Sighting down this bond, the central atom's parent is placed on the triangle at an arbitrary position. Children are then iterated counterclockwise or clockwise relative to it.

@@

\begin{table}
\caption{Balsa's Character Set.}
\centering
\begin{tabular}{c c c c c c}
    \hline
    Code & Character & Code & Character & Code & Character      \\
    \hline
    35   & \ttt{\#}  & 67   & \ttt{C}   & 92   & \ttt{\char`\\} \\
    37   & \ttt{\%}  & 68   & \ttt{D}   & 93   & \ttt{]}        \\
    40   & \ttt{(}   & 69   & \ttt{E}   & 97   & \ttt{a}        \\
    41   & \ttt{)}   & 70   & \ttt{F}   & 98   & \ttt{b}        \\
    42   & \ttt{*}   & 71   & \ttt{G}   & 99   & \ttt{c}        \\
    43   & \ttt{+}   & 72   & \ttt{H}   & 100  & \ttt{d}        \\
    45   & \ttt{-}   & 73   & \ttt{I}   & 101  & \ttt{e}        \\
    46   & \ttt{.}   & 75   & \ttt{K}   & 102  & \ttt{f}        \\
    47   & \ttt{/}   & 76   & \ttt{L}   & 103  & \ttt{g}        \\
    48   & \ttt{0}   & 77   & \ttt{M}   & 104  & \ttt{h}        \\
    49   & \ttt{1}   & 78   & \ttt{N}   & 105  & \ttt{i}        \\
    50   & \ttt{2}   & 79   & \ttt{O}   & 107  & \ttt{k}        \\
    51   & \ttt{3}   & 80   & \ttt{P}   & 108  & \ttt{l}        \\
    52   & \ttt{4}   & 82   & \ttt{R}   & 109  & \ttt{m}        \\
    53   & \ttt{5}   & 83   & \ttt{S}   & 110   & \ttt{n}       \\
    54   & \ttt{6}   & 84   & \ttt{T}   & 111   & \ttt{o}       \\
    55   & \ttt{7}   & 85   & \ttt{U}   & 112   & \ttt{p}       \\
    56   & \ttt{8}   & 86   & \ttt{V}   & 114   & \ttt{r}       \\
    57   & \ttt{9}   & 87   & \ttt{W}   & 115   & \ttt{s}       \\
    61   & \ttt{=}   & 88   & \ttt{X}   & 116   & \ttt{t}       \\
    64   & \ttt{@}   & 89   & \ttt{Y}   & 117   & \ttt{u}       \\
    65   & \ttt{A}   & 90   & \ttt{Z}   & 121   & \ttt{y}       \\
    66   & \ttt{B}   & 91   & \ttt{[}   &       &               \\
    \hline
\end{tabular}
\label{table:character-set}
\end{table}

@@

Balsa's formal grammar is presented as a series of \textit{production rules} (aka \enquote{productions}). A production defines a transformation allowed under the grammar. These transformations collectively define the set of valid Balsa strings. Productions can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string.

@@

Atoms carry most of the information in a Balsa string. The \ttt{atom} non-terminal can be encoded using one of four productions.

@@

The first production, \ttt{star} (\enquote{start atom}), produces one terminal value, the asterisk character (\ttt{*}). This \enquote{star atom} represents an atom in which every attribute assumes its default value.

@@

The third atomic production, \ttt{selection} (\enquote{selection}) is a non-terminal selected from the list: \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding capitalized atom symbol to the \ttt{element} attribute, sets the \ttt{selected} attribute to \ttt{true}, and sets the \ttt{hydrogens} attribute to \ttt{Implicit}. All other atomic attributes retain their default values.

@@

Given the large number of variants within the \ttt{element} non-terminal, the following production only defines the first several. For a complete list of variants, see the full grammar in the Supporting Information.

@@

Children are added to an atom through the \ttt{sequence} non-terminal (\enquote{sequence}). A sequence starts with a required \ttt{atom} non-terminal. If an allowed non-terminal does not follow, the corresponding atom will have no children. Allowed non-terminals will be one of: \ttt{union}; \ttt{branch}; or \ttt{split}.

\begin{lstlisting}
sequence ::= atom ( union | branch | split )*
\end{lstlisting}

@@

The \ttt{bond} non-terminal supports five variants (\enquote{-,} \enquote{=,} \enquote{\#,} \enquote{/,} and \enquote{\textbackslash}). The first three (\enquote{-,} \enquote{=,} and \enquote{\#}) set the \ttt{order} attribute of a bond to one, two, or three, respectively. The last two terminals, \enquote{/} and \enquote{\textbackslash,} set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} or \ttt{Down}, respectively.

@@

\begin{table}
    \caption{Stack and hub for reading the string \enquote{C(N*)O}.}
    \centering
    \begin{tabular}{c c l c c}
        \hline
        Step & Character   & Action                          & Hub & Stack \\
        \hline
        1    & \enquote{C} & create atom                     & C   &       \\
        2    & \enquote{(} & push next atom                  & C   &       \\
        3    & \enquote{N} & create atom, push atom          & N   &  N    \\
        4    & \enquote{*} & create atom                     & *   &  N    \\
        5    & \enquote{)} & pop atom, update hub            & N   &       \\
        6    & \enquote{O} & create atom                     & O   &       \\
        \hline
    \end{tabular}
    \label{table:stack-for-branches}
\end{table}

Additional modifications allow bridge bonds to be tested for \textit{compatibility} (Figure~\ref{fig:bridge-bond-compatibility}). Compatibility is tested by comparing the attributes of each bridge bond. This test has three parts:

@@

A workflow modification allows bridge bonds to be compared for compatibility. A 100-element array is used to maintain state. The first occurrence of a bridge causes the bridge bond's attributes to be copied to the array at the bridge index. The second occurrence of a bridge causes the entry to be removed. The attributes stored in the entry are then compared with those of the current bridge bond. If the two sets of attributes are incompatible, an error must be reported.

@@

If a selected atom bears a non-zero \ttt{charge} attribute, subvalence is computed using the isoelectronic element's default valences. For example, a selected nitrogen atom with a charge of +1 would use the default valence for carbon. A selected phosphorous atom with a charge of -1 would use the default valences of sulfur. And so on. If no default valences are found in this way (e.g., \enquote{[c+2]}), a reader must report an error. Writers must not encode such atoms.

@@

Regardless the input data structure's form, it must be traversable in \textit{depth-first} order (Figure~\ref{fig:depth-first-traversal}). A depth-first traversal operates over a set of nodes bound by a set of connectivity relationships, which are typically edges. Traversal proceeds by successively replacing each node as the center of focus, or root. Each new root is chosen from the untraversed neighbors of the current root. Eventually all nodes are processed, ending the traversal.

@@

A writer intercepts the depth-first traversal of an input data structure to write an output string compatible with the formal grammar. There are no requirements around style. For example, it's equally valid to represent the carbon atom of methane using either implicit or virtual hydrogens (e.g., \enquote{C} or \enquote{[CH4]}). Single bonds may be elided or not, all other things being equal. Similarly, selection may or may not be used. Although an organization may seek to standardize certain styles of string output, a reader must process any syntactically and semantically valid string.

As described previously, the \ttt{branch} non-terminal encodes branches. A useful tool in this regard is a stack (Table~\ref{table:stack-for-writing-string}). A writer begins by pushing the current branch onto the stack and extending it. When a new branch is encountered, it is pushed to the stack and extended. When the branch terminates, the current branch is popped and its contents are appended to the stack's new top item.

@@

Molecular trees offer some important advantages, but they are used less often than molecular graphs. Sooner or later, interconversions will be required. Readers will need to transform a tree into a graph, and writers will need to transform a graph into a tree. Although there are only three main differences between molecular trees and molecular graphs, these differences can lead to complex reconciliations during transformation. Understanding them is the key to lossless conversion.

@@

Nevertheless, several SMILES software implementations have been released as noted previously. Each one represents its own unique set of decisions about how to reduce the published broad SMILES descriptions to a specification detailed enough for software development. Some implementations have themselves been reverse-engineered for clues. The Daylight implementation in particular is sometimes consider a de facto SMILES specification. Unfortunately, access to this implementation is restricted by commercial licensing, as is the case with several SMILES implementations. Moreover, reverse engineering is unlikely to address every question because of the vast search space defined by SMILES. Although some implementations are released under Open Source licenses, they are neither authoritative nor unanimous in their interpretations.

A publicly-available and precise subset of SMILES would offer a new way to address the problem. If sufficiently constrained, the subset should be interpretable by all existing software without modification. Moreover, most strings written by existing software would be interpretable by software based on the language subset. Because its description would be both public and precise, the language subset could serve as the basis for many implementations with strong consensus around both syntax and semantics.

@@

Accordingly, Balsa's base language is \enquote{ProtoSMILES,} a SMILES proto-language. ProtoSMILES was identified through a process of elimination that started with every SMILES document ever published and every SMILES software implementation. The field was narrowed through two restrictions: (1) the source must be authoritative; and (2) the source must be publicly accessible. \enquote{Authoritative} in this context means that the source was created by either Weininger himself or Daylight, who alone can speak most authoritatively about what SMILES is. The requirement for public accessibility ensures that claims of compatibility can be tested without limit. The following documents were identified as candidate ProtoSMILES sources:

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency (\enquote{The Report})~\cite{anderson:1987}.
    \item Weininger's original publication (herein \enquote{The Article})~\cite{weininger:1988}.
    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Chapter})~\cite{weininger:2003}.
    \item An online manual maintained by Daylight (\enquote{The Manual}).
    \item The Daylight's toolkit, which implements a SMILES reader and writer (\enquote{The Toolkit})~\cite{daylightToolkit}.
\end{itemize}

>>>>

\begin{abstract}
Despite its widespread use, Simplified Molecular Input Line Entry System (SMILES) remains underspecified. The lack of a detailed specification encourages improvisation by software developers, complicates data standardization efforts, and undermines extension development. Balsa, a reformulation of SMILES, addresses these problems. A formal, machine-readable grammar defines Balsa's syntax. Semantics are described at three levels of resolution: constitution; configuration; and conformation. To estimate the compatibility of Balsa strings with SMILES software, all differences between the two languages were enumerated. To the extent that SMILES documentation is unambiguous, every difference was consistent with Balsa being a language subset.
\end{abstract}

@@

Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988~\cite{weininger:1988}. As a line notation~\cite{wiswesser:1968}, SMILES represents molecules as single line character sequences, or strings. SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel~\cite{oBoyle:2011}; RDKit~\cite{rdkit}; Chemistry Development Kit~\cite{steinbeck:2003}; JChem Base~\cite{jchemBase}; the Daylight Toolkit~\cite{daylightToolkit}; and OEChem TK~\cite{oetk}. SMILES encodings can be found in many public-facing databases, including: PubChem~\cite{kim:2016}; ChEBI~\cite{degtyarenko:2007}; ZINC~\cite{irwin:2005}; ChEMBL~\cite{gaulton:2012}; and Wikipedia~\cite{wikipediaCAS}. Raw SMILES strings have been used extensively in machine learning applications~\cite{sousa:2021}. Algorithmic selection of a single SMILES encoding has been used for molecular identification~\cite{weininger:1989,oBoyle:2012}. SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES~\cite{hanson:2016}, CurleySMILES~\cite{drefahl:2011}, BigSMILES~\cite{lin:2019}, and CXSMILES~\cite{cxsmiles}.

@@

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's original description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2003}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. (\enquote{Daylight}), recapitulates published material~\cite{daylightTheory}. The lack of other authoritative sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

@@

More broadly, this situation undermines ongoing data integrity efforts underway in other scientific disciplines. As an example, the FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. The lack of precision in existing SMILES documentation means that chemical data sets based on SMILES will suffer in all four areas.

@@

Multiple molecules are supported through \textit{gaps}. A gap is an edge that defines no bonding relationship. Gaps can be used whenever two or more disconnected molecules need to be encoded within the same molecular tree. Specific applications include ionic bonding and non-ionic complexes. However, no bonding relationship whatsoever is implied by the presence of a gap. In particular, the zero-order bond interpretation is excluded \cite{clark:2011}.

@@

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen atom and its bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. Not all atoms are eligible for this treatment (see: \nameref{computing-implicit-hydrogen-count}).

It is sometimes convenient to interconvert virtual and atomic hydrogens. The transformation of an atomic hydrogen into a virtual hydrogen is called \textit{reification}. The reverse process is called \textit{virtualization}.

@@

\begin{table}
\caption{\ttt{AtomParity} Variants.}
\centering
\begin{tabular}{l l}
    \hline
    Variant & Description  \\
    \hline
    \ttt{Clockwise} & Configuration runs with node order. \\
    \ttt{Counterclockwise} & Configuration runs against node order. \\
    \hline
\end{tabular}
\label{table:atom-parity-variants}
\end{table}

The \ttt{edges} (\enquote{edges}) attribute is a possibly empty ordered list of edges associated with an atom. As noted in the previous section, an edge may represent either a bond or a gap (Table~\ref{table:edge-variants}). The mere presence of an edge does not guarantee a bonding relationship.

@@

A bond is composed of four attributes (Table~\ref{table:bond-attributes}). The \ttt{order} attribute (\enquote{order}) represents the concept of formal bond order, or the number of bonding electrons divided by two. A bond's order may only assume the values one, two, or three. When set to \ttt{true}, the boolean \ttt{elided} attribute (\enquote{elided}) allows a bond to be omitted during serialization, and to participate in extended bonding. The \ttt{direction} attribute (\enquote{direction}) is used with other information to conformationally restrict a bond. The two allowed variants for \ttt{Direction}, \ttt{Up} and \ttt{Down} (Table~\ref{table:bond-direction-variants}), will be described in detail later (see: \nameref{conformation}). The \ttt{target} attribute identifies the child node at which the bond is directed (Table~\ref{table:target-variants}).

\begin{table}
\caption{Bond Attributes.}
\centering
\begin{tabular}{l p{3.6cm} l l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{order} & Formal bond order & \ttt{1..4} & 1 \\
    \ttt{elided} & Elision status & \ttt{?} & \ttt{false} \\
    \ttt{direction} & Partial conformational parity & \ttt{\{Direction\}} & \ttt{None} \\
    \ttt{target} & An atom or bridge node & \ttt{Target} & - \\
    \hline
\end{tabular}
\label{table:bond-attributes}
\end{table}

@@

The atom and bond attributes defined here will be referenced throughout this paper as a way to connect syntax and semantics. Implementations may, of course, use any suitable internal data model. They must, however, ensure that the model used is consistent with the one provided here.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME, Figure~\ref{fig:dime}). DIME results from limitations to the valence bond model causing two representations to encode what would be considered the same molecule. When the two forms differ only in double bond placement, the terms \enquote{resonance} or \enquote{aromaticity} are sometimes used. These terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

Selection is nothing more than an alternative to explicit bond order encoding. No other properties whatsoever are conferred to selected atoms. The members of a DS may be assigned the labels \enquote{aromatic,} \enquote{antiaromatic,} \enquote{non-aromatic,} or any number of other designations by downstream applications. Two encodings differing only in the presence of a DS should be considered equivalent. Balsa imposes no rules around whether or not a selectable atom must be selected, although users may do so voluntarily.

@@

To support implicit hydrogen counting, Balsa uses the concept of \textit{valence}. Valence is a non-negative integer tally computed as the sum of bond orders at a given atom. Single and elided bonds contribute one to the tally, double bonds two, and triple bonds three. Each virtual hydrogen contributes one. For example, the valence of a methyl carbon having three virtual hydrogens and a hydrogen neighbor is four. The valence of an acetaldehyde oxygen atom is two. And so on.

@@

Restricted rotation about a bond leads to conformational isomers (\enquote{conformers}). When the bond in question is a double bond, two conformers result. These can be distinguished through \textit{partial parity bonds} (PPBs). As the name implies, a PPB encodes some of the parity information characterizing a conformationally restricted double bond. Reconstruction of the full parity requires the double bond itself and at least one flanking PPB at each terminal.

@@

The configuration about an atom can be specified through its \ttt{parity} attribute. This attribute, together with a general convention, determine the relative three-dimensional positioning of substituents about a central atom (Figure~\ref{fig:atom-parity}). First, the parent of the central atom is identified. Then the frame of reference is shifted by sighting down the bond from the central atom's parent to the central atom. An equilateral triangle is then inscribed about the central atom. The bonds from the central atom are then iterated in order. If the parity of the central atom equals \ttt{Clockwise}, children placed around the triangle in clockwise order. Otherwise the children are placed in counterclockwise order.

@@

The convention is modified for atoms bearing one virtual hydrogen and three atomic substituents (Figure~\ref{fig:atom-parity-suppressed-hydrogen}). The implied bond to the virtual hydrogen is considered to precede all other children.

@@

\begin{table}
\caption{Balsa's US-ASCII Derived Character Set.}
\centering
\begin{tabular}{c c c c c c}
    \hline
    Code & Character & Code & Character & Code & Character      \\
    \hline
    35   & \ttt{\#}  & 67   & \ttt{C}   & 92   & \ttt{\char`\\} \\
    37   & \ttt{\%}  & 68   & \ttt{D}   & 93   & \ttt{]}        \\
    40   & \ttt{(}   & 69   & \ttt{E}   & 97   & \ttt{a}        \\
    41   & \ttt{)}   & 70   & \ttt{F}   & 98   & \ttt{b}        \\
    42   & \ttt{*}   & 71   & \ttt{G}   & 99   & \ttt{c}        \\
    43   & \ttt{+}   & 72   & \ttt{H}   & 100  & \ttt{d}        \\
    45   & \ttt{-}   & 73   & \ttt{I}   & 101  & \ttt{e}        \\
    46   & \ttt{.}   & 75   & \ttt{K}   & 102  & \ttt{f}        \\
    47   & \ttt{/}   & 76   & \ttt{L}   & 103  & \ttt{g}        \\
    48   & \ttt{0}   & 77   & \ttt{M}   & 104  & \ttt{h}        \\
    49   & \ttt{1}   & 78   & \ttt{N}   & 105  & \ttt{i}        \\
    50   & \ttt{2}   & 79   & \ttt{O}   & 107  & \ttt{k}        \\
    51   & \ttt{3}   & 80   & \ttt{P}   & 108  & \ttt{l}        \\
    52   & \ttt{4}   & 82   & \ttt{R}   & 109  & \ttt{m}        \\
    53   & \ttt{5}   & 83   & \ttt{S}   & 110   & \ttt{n}       \\
    54   & \ttt{6}   & 84   & \ttt{T}   & 111   & \ttt{o}       \\
    55   & \ttt{7}   & 85   & \ttt{U}   & 112   & \ttt{p}       \\
    56   & \ttt{8}   & 86   & \ttt{V}   & 114   & \ttt{r}       \\
    57   & \ttt{9}   & 87   & \ttt{W}   & 115   & \ttt{s}       \\
    61   & \ttt{=}   & 88   & \ttt{X}   & 116   & \ttt{t}       \\
    64   & \ttt{@}   & 89   & \ttt{Y}   & 117   & \ttt{u}       \\
    65   & \ttt{A}   & 90   & \ttt{Z}   & 121   & \ttt{y}       \\
    66   & \ttt{B}   & 91   & \ttt{[}   &       &               \\
    \hline
\end{tabular}
\label{table:character-set}
\end{table}

@@

Balsa's formal grammar is presented as a series of \textit{production rules} (\enquote{productions}). A production defines a transformation allowed under the grammar. These transformations collectively define the set of valid Balsa strings. Productions can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string.

@@

Atoms carry most of the information in a string. The \ttt{atom} non-terminal can be encoded using one of four productions.

@@

The first production, \ttt{star} (\enquote{star atom}), is comprised of one terminal value, the asterisk character (\ttt{*}). This \enquote{star atom} represents an atom in which every attribute assumes its default value.

@@

The third atomic production, \ttt{selection} (\enquote{selection}), is a non-terminal selected from the list: \enquote{b}; \enquote{c}; \enquote{n}; \enquote{o}; \enquote{p}; and \enquote{s.} An atom encoded in this way (\enquote{selected shortcut atom}) assigns the corresponding capitalized atom symbol to the \ttt{element} attribute, sets the \ttt{selected} attribute to \ttt{true}, and sets the \ttt{hydrogens} attribute to \ttt{Implicit}. All other atomic attributes retain their default values.

@@

Given the large number of variants within the \ttt{element} non-terminal, the following production only defines the first several. For a complete list of variants, see the full grammar in the Supporting Material.

@@

Children are added to an atom through the \ttt{sequence} non-terminal (\enquote{sequence}). A sequence starts with a required \ttt{atom} non-terminal. If an allowed non-terminal does not follow, the corresponding atom will have no children. Allowed non-terminals will be one of: \ttt{union}; \ttt{branch}; or \ttt{gap}.

\begin{lstlisting}
sequence ::= atom ( union | branch | gap )*
\end{lstlisting}

@@

The \ttt{bond} non-terminal supports five variants (\enquote{-,} \enquote{=,} \enquote{\#,} \enquote{/,} and \enquote{\textbackslash}). The first three (\enquote{-,} \enquote{=,} and \enquote{\#}) set the \ttt{order} attribute of a bond to one, two, or three, respectively. The last two terminals (\enquote{/} and \enquote{\textbackslash}) set the \ttt{order} attribute to 1 while also setting the \ttt{direction} attribute to \ttt{Up} or \ttt{Down}, respectively.

@@

\begin{table}
    \caption{Stack and hub for reading the string \enquote{C(N*)O}.}
    \centering
    \begin{tabular}{c c l c c}
        \hline
        Step & Character   & Action                          & Hub & Stack \\
        \hline
        1    & \enquote{C} & create atom                     & C   &       \\
        2    & \enquote{(} & push hub to stack               & C   &  C    \\
        3    & \enquote{N} & create, connect atom            & N   &  C    \\
        4    & \enquote{*} & create, connect atom            & *   &  C    \\
        5    & \enquote{)} & pop atom, update hub            & C   &       \\
        6    & \enquote{O} & create, connect atom            & O   &       \\
        \hline
    \end{tabular}
    \label{table:stack-for-branches}
\end{table}

Bridge bonds must be tested for \textit{compatibility} (Figure~\ref{fig:bridge-bond-compatibility}), a task complicated by the presence of PPBs. A three-part test can be used:

@@

These tests can be implemented using a 100-element array. The first occurrence of a bridge causes the bridge bond's attributes to be copied to the array at the bridge index. The second occurrence of a bridge causes the entry to be removed. The attributes stored in the entry are then compared with those of the current bridge bond. If the two sets of attributes are incompatible, an error is reported.

@@

An atom to be considered for pruning may have a non-zero charge. If so, subvalence is computed using the isoelectronic element's default valences. For example, a selected nitrogen atom with a charge of +1 would use the default valence for carbon. A selected phosphorous atom with a charge of -1 would use the default valences of sulfur. And so on. If no default valences are found in this way (e.g., \enquote{[c+2]}), a reader must report an error. Writers must not encode such atoms.

@@

Regardless the input data structure's form, it must be traversable in \textit{depth-first} order (Figure~\ref{fig:depth-first-traversal}). A depth-first traversal operates over a set of nodes associated in pairwise fashion by a set of connectivity relationships, which are typically edges. Traversal proceeds by successively replacing each node as the center of focus, or root. Each new root is chosen from the untraversed neighbors of the current root. Eventually all nodes are processed, ending the traversal.

@@

A writer intercepts the depth-first traversal of an input data structure to write an output string compatible with the formal grammar. There are no requirements around style. For example, it's equally valid to represent the carbon atom of methane using either implicit or virtual hydrogens (e.g., \enquote{C} or \enquote{[CH4]}). Single bonds may be elided or not, all other things being equal. Similarly, selection may or may not be used. Although an organization may seek to standardize certain styles of string output, a reader must process any string that is syntactically and semantically valid.

As described previously, the \ttt{branch} non-terminal encodes branches. A useful tool for encoding branches is a stack (Table~\ref{table:stack-for-writing-string}). A writer begins by pushing the current branch onto the stack and extending it. When a new branch is encountered, it is pushed to the stack and extended. When the branch terminates, the current branch is popped and its contents are appended to the stack's new top item.

@@

Molecular trees offer some important benefits, but they are used less often than molecular graphs. Sooner or later, interconversions will be required. Readers will need to transform a tree into a graph, and writers will need to transform a graph into a tree. Although there are only three main differences between molecular trees and molecular graphs, these differences can lead to complex reconciliations during transformation. Understanding them is the key to lossless conversion.

@@

Nevertheless, several SMILES software implementations have been released. Each one represents a unique set of decisions about how to reduce the published broad SMILES descriptions to a specification detailed enough for software development. Some implementations have themselves been reverse-engineered for clues. The Daylight implementation in particular is sometimes considered a de facto SMILES specification. Unfortunately, access to this implementation is restricted by commercial licensing, as is the case with several SMILES implementations. Moreover, reverse engineering is unlikely to address every question because of the vast search space defined by SMILES. Although some implementations are released under Open Source licenses, they are neither authoritative nor unanimous in their interpretations.

A publicly-available and precise subset of SMILES would offer a new way to address the problem. Such a subset should be interpretable by all existing software without modification. Moreover, most strings written by existing software would be interpretable by software based on the language subset. Because its description would be both public and precise, the language subset could serve as the basis for many implementations, with strong consensus around both syntax and semantics.

@@

Accordingly, Balsa's base language is \enquote{ProtoSMILES,} a SMILES proto-language. ProtoSMILES was identified through a process of elimination that started with every SMILES document ever published and every SMILES software implementation. The field was narrowed through two restrictions: (1) the source must be authoritative; and (2) the source must be publicly accessible. \enquote{Authoritative} in this context means that the source was created by either Weininger himself or Daylight, who alone can speak most authoritatively about what SMILES is. The requirement for public accessibility ensures that claims of compatibility can be tested. The following documents were identified as candidate ProtoSMILES sources:

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency (\enquote{The Report})~\cite{anderson:1987}.
    \item Weininger's original publication (herein \enquote{The Article})~\cite{weininger:1988}.
    \item Weininger's 2003 book chapter, revised in 2008 (\enquote{The Chapter})~\cite{weininger:2003}.
    \item An online manual maintained by Daylight (\enquote{The Manual})~\cite{daylightTheory}.
    \item The Daylight's toolkit, which implements a SMILES reader and writer (\enquote{The Toolkit})~\cite{daylightToolkit}.
\end{itemize}
